---
output: 
  stevetemplates::syllabus:
    latex_engine: xelatex
    dev: cairo_pdf
geometry: margin=1in

title: "Principles of Data Science"
author: "PSYC 259"
date: "Winter 2022"

email: "franchak@ucr.edu"
web: "padlab.ucr.edu"
officehours: "By appointment"
office: "https://ucr.zoom.us/my/johnfranchak"
classroom: "Psychology 1205"
classhours: "Mondays 1:00pm-3:50pm"

fontfamily: mathpazo
fontsize: 11pt
header-includes:
   - \linespread{1.05}
---


```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(fig.path='figs/')
knitr::opts_chunk$set(cache.path='_cache/')
knitr::opts_chunk$set(warning=F, message=F)
```

```{r dates_bib, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, dev='pdf')
wed <- as.Date("2021-01-03")

advdate <- function(obj, adv) {
 tmon <- obj + 7*(adv-1)
 tfri <- obj + 4 + 7*(adv-1)
 tmon <- format(tmon, format="%m/%d")
 tfri <- format(tfri, format="%m/%d")
 zadv <- sprintf("%02d", adv)
 #tmp <- paste("Week ",zadv,sep='',", ", tmon," - ",tfri)
  tmp <- paste("Week ",zadv,sep='',", ", tmon)
 return(tmp)
}
```

# TA Information

* Jacob Elder 
* Office Hours: TBD 
* E-mail: jacob.elder@email.ucr.edu 
* Zoom: https://ucr.zoom.us/my/jelder 


# Course Description

Most quantitative courses (importantly) focus on the final steps of data analysis---conducting and understanding statistical tests. However, much of the work in data science is taking raw data, often from multiple, incompatible sources, and processing those data into a usable form. This course will emphasize the importance of robust, documented, and automated workflows for processing data to save time, reduce errors, improve reproducibility, and facilitate collaboration among multiple researchers. We will also spend time on data visualization and communication---an important part of creating, checking, and collaborating on data workflows. We will use the R programming language, Github, and Rmarkdown to work through examples, but the focus is on concepts/best practices that can be applied to any software or programming language. The course is open to students who have little programming experience or experience with R. The goal is for students at all levels of programming experience to set goals to improve their data science skills. 

# Course Objectives

The goals of this course are for you to critically analyze and improve your data analysis workflows. Implementing robust, automated procedures for handling data will allow you to:

- Foster open science through increased transparency, reproducibility, and easier data sharing
- Increase the fidelity of your data and analyses by detecting and preventing errors
- Better understand and communicate about your data
- Save time by preventing errors, automating tasks, and reusing code
- Facilitate collaboration with organized and documented workflows

# Course Materials

- The course [Github page](https://github.com/orgs/psych-259-data-science-2022/repositories) has links to project files, readings, and the syllabus
- Readings from *R for Data Science* are available [online](https://r4ds.had.co.nz/index.html).
- PDFs of other course readings referenced below are available on Github in the syllabus folder.

## Other helpful resources
- RStudio [cheatsheets](https://rstudio.com/resources/cheatsheets/) for base R, data import, ggplot, R Markdown, and other packages.
- [STAT 545](https://stat545.com/index.html), another good, free textbook on data wrangling and exploration
- [Happy Git with R](https://happygitwithr.com), for all of your Github-Git-R-RStudio needs
- [*R Markdown: The Definitive Guide*](https://bookdown.org/yihui/rmarkdown/)
- [*The Science of Visual Data Communication* article](https://doi.org/10.1177%2F15291006211051956)

# Course Policies

Please follow the course [Github page](https://github.com/orgs/psych-259-data-science-2022/repositories) for updates to the syllabus and materials. I will communicate changes through email and/or Canvas announcements. 

## Assignments and Grading

Grading for the class is S/NC. Your grade will be based on the following:

```{=latex}
\begin{tabular}{lc}
Component & Weight \\
\hline
Participation           & 20\% \\
Weekly assignments      & 20\% \\
Workflow critique       & 20\% \\
Final project           & 40\% \\
\end{tabular}
```

### Participation 

You are expected to attend each class and participate in class discussions. Readings should be completed prior to class so that you can contribute to discussions. 

### Weekly Assignments 

I will assign short practical assignments after most weeks of class. These exercises will develop your skills in R programming and other concepts that we discuss. You are allowed to work in groups of 2-3 students to complete these exercises if you would like, but each student should turn in their own assignment. Please indicate in the top of your assignment the names of the students you worked with. Some class time will be given so that you can start work on the assignment with the help of the instructor/TA. Office hours are also available for help on assignments. 

### Workflow Self-Critique 

You should choose a current or past data analysis project that you have worked with (or one from your lab if you are a newer student). In a short paper (3-4 pages), you will first describe the end-to-end workflow of your data. What are sources of raw data? How are those sources combined and/or processed? What research personnel, computing resources, software, and hardware devices are involved? What is the end product needed for statistical analysis? Next, you will take a critical eye to your workflow and identify 1) Where are errors most likely to occur, 2) What time-consuming steps could be automated, and 3) how your workflow could be made more transparent and reproducible. You will briefly present your workflow critique in class. Each student will work individually on this assignment.

### Final Project 

In your final project, you will improve the data workflow that you chose using skills learned in this class. Your final project should be shared with the instructor and TA through an online repository (such as Github, OSF, or Code Ocean) and allow your end-to-end data workflow to be reproduced (e.g., include raw data files, functions that implement processing steps, etc.). You can use whatever programming languages are necessary (it doesn't need to just be in R), but you should consult with the instructor if R will not be used to ensure that the instructor can run your code (or alternatively, that you demonstrate your workflow to the instructor). Your project should be a report, either to demonstrate exploratory analyses or to communicate the results, that is written in R Markdown and contains visualizations written in R. 

## Attendance

In the event of absences due to illness or family obligations, please contact the instructor immediately so that we can make arrangements. Everyone is in a different situation, so I will always work with you to figure out a way to handle disruptions to learning. I want everyone in this class to succeed!

\newpage

# Class Schedule

Readings should be completed prior to each class. 

## `r advdate(wed, 1)`: Course Goals and Setup
- Goals of the class
- Logistics
- Workflow examples
- Software setup
- Introduction to Github

*R4DS:* [Introduction](https://r4ds.had.co.nz/introduction.html) and [Workflow Basics](https://r4ds.had.co.nz/workflow-basics.html)

*GitHub Project Links:*

- [Instructions to set up R, RStudio, git, and Github](https://github.com/psych-259-data-science-2022/259-syllabus-readings-slides/blob/master/software_setup.md)

## `r advdate(wed, 2)`: File Organization and Workflow
- File organization 
- Version control
- *SKILLS:* R language basics (variables, assignment, comments, calling functions), file import/export

*R4DS:* [Scripts](https://r4ds.had.co.nz/workflow-scripts.html), [Projects](https://r4ds.had.co.nz/workflow-projects.html),  [Tibbles](https://r4ds.had.co.nz/tibbles.html), and [Data import](https://r4ds.had.co.nz/data-import.html)

*GitHub Project Links:*

- [R language basics](TBD)
- [File import](TBD)
- [Import homework](TBD)

##  `r advdate(wed, 3)`: NO CLASS - MLK DAY

##  `r advdate(wed, 4)`: Data Transformations
- Basic data types (numbers, strings)
- Variable naming
- *SKILLS:* dplyr data wrangling, logical statements, strings

*R4DS:* [Data transformation](https://r4ds.had.co.nz/transform.html) and [Strings](https://r4ds.had.co.nz/strings.html)

*OTHER READINGS:* Mistakes article (Kovacs et al., 2021)

*GitHub Project Links:*

- [Data wrangling](TBD)
- [Data wrangling homework](TBD)

##  `r advdate(wed, 5)`: Data Structure
- Tidy data (observations/cases)
- More advanced data types (factors, dates, vectors, lists)
- *SKILLS:* More data wrangling, factors, dates

*R4DS:* [Tidy data](https://r4ds.had.co.nz/tidy-data.html), [Factors](https://r4ds.had.co.nz/factors.html), and [Dates](https://r4ds.had.co.nz/dates-and-times.html)

*GitHub Project Links:*

- [Data shaping](TBD)
- [Homework](TBD)

##  `r advdate(wed, 6)`: Automation: Functions and Iteration
- Finding new packages/APIs vs. writing your own functions
- Automating your analyses
- Writing more efficient code
- *SKILLS:* Writing custom functions, iteration (map, for loops)

*R4DS:* [Functions](https://r4ds.had.co.nz/functions.html), [Vectors](https://r4ds.had.co.nz/vectors.html) and [Iteration](https://r4ds.had.co.nz/iteration.html)

*OTHER READINGS:* Minimizing mistakes article (Rouder et al., 2019)

*GitHub Project Links:*

- [Automation](TBD)
- [Automation homework](TBD)

##  `r advdate(wed, 7)`: Exploratory Data Analysis
- **WORKFLOW PRESENTATIONS**
- Data validation
- Automating visualizations 
- *SKILLS:* Basic visualizations with ggplot2

*R4DS:* [Data visualization](https://r4ds.had.co.nz/data-visualisation.html) and [Exporatory data analysis](https://r4ds.had.co.nz/exploratory-data-analysis.html)

*GitHub Project Links:*

- [Visualization and EDA](TBD)
- [Visualization and EDA homework](TBD)

##  `r advdate(wed, 8)`: NO CLASS: PRESIDENTS' DAY

##  `r advdate(wed, 9)`: Data Sharing and Reproducibility
- Reuse-minded project management
- Reproducible reports
- Preserving programming environment and analyses
- *SKILLS:* R Markdown, package control

*R4DS:* [R Markdown](https://r4ds.had.co.nz/r-markdown.html)

*OTHER READINGS:* Transparency in psychological science (Klein et al., 2018)

*GitHub Project Links:*

- [RMarkdown](TBD)

##  `r advdate(wed, 10)`: Communication
- Communicating through graphical styles
- Interactive plots for data exploration
- *SKILLS:* ggplot and extensions

*R4DS:* [Graphics for communication](https://r4ds.had.co.nz/graphics-for-communication.html)

*OTHER READING:* [Chartjunk](https://www.edwardtufte.com/bboard/q-and-a-fetch-msg?msg_id=00040Z) from (Tufte 1990; 2001; 2006)

*GitHub Project Links:*

- [Visualization](TBD)
